---
title: "ggplot data visualisation"
output: html_notebook
---

In this question, we are going to introduce the use of ggplot to visualise data for soil-transmitted helminths collected via a cross-sectional study that was conducted in nine villages across three districts of Meghalaya: Ri Bhoi, Eastern West Khasi Hills, and Jaintia Hills.

Load in the data and clean it to remove NAs in the `overall_sth_microscopy` and `age` variables. Also remove observations for individuals aged 60 or above or where the age is zero.
```{r}
library(tidyverse)
filename <- "../../data/sth_data.csv"
df <- read.csv(filename)
  
# remove NAs in age and seropositivity
df <- df %>% 
  filter(!is.na(overall_sth_microscopy)) %>% 
  filter(!is.na(age)) %>% 
  filter(age < 60) %>% 
  filter(age > 0)
```

Load the tidyverse package (really, tidyverse is a collection of packages).
```{r}
library(tidyverse)
```

0. Use ggplot to produce a histogram which visualises the distribution of ages in the dataset.
```{r}
df %>% 
  ggplot(aes(x=age)) +
  geom_histogram()
```
0a. Create histograms of age by sex, with different colours for the genders.
```{r}
df %>% 
  ggplot(aes(x=age, fill=gender)) +
  geom_histogram(position = "identity", alpha=0.4)
```


0b. Create histograms of age by sex, with separate panels for males and females.
```{r}
df %>% 
  ggplot(aes(x=age)) +
  geom_histogram() +
  facet_wrap(~gender)
```
0c. Produce boxplots of the age distribution by sex.
```{r}
df %>% 
  ggplot(aes(x=gender, y=age)) +
  geom_boxplot()
```

Od. Produce violin plots of the age distribution by sex.
```{r}
df %>% 
  ggplot(aes(x=gender, y=age)) +
  geom_violin()
```
Od. Produce violin plots of the age distribution by sex.

1. Calculate the proportion positive by village.
```{r}
df_villages <- df %>% 
  group_by(village) %>% 
  summarise(
    prop_positive=mean(overall_sth_microscopy)
  )
```


2. Use ggplot to produce a point plot of the proportion testing positive via microscopy (i.e. the `overall_sth_microscopy` by variable) by village.

Hint: put the village name on the horizontal axis.

```{r}
df_villages %>% 
  ggplot(aes(x=village, y=prop_positive)) +
  geom_point()
```
3. Now we are going to tidy up the above plot by relabeling the vertical axis so that it produces %s. To do so, add the following to your above plot command:

` + scale_y_continuous(labels=scales::percent)`

```{r}
df_villages %>% 
  ggplot(aes(x=village, y=prop_positive)) +
  geom_point() +
  scale_y_continuous(labels=scales::percent)
```
4. Change the horizontal axis label to be "Village" and the vertical axis to be "Prevalence, %".
```{r}
df_villages %>% 
  ggplot(aes(x=village, y=prop_positive)) +
  geom_point() +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %")
```
5. We are now going to change the plot theme to remove the grey background to it. To do so, add the following to your plot:

`+ theme_classic()`

```{r}
df_villages %>% 
  ggplot(aes(x=village, y=prop_positive)) +
  geom_point() +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic()
```
6. Flip the order of your horizontal and vertical axes by trying out the `+ coord_flip()` command.

```{r}
df_villages %>% 
  ggplot(aes(x=village, y=prop_positive)) +
  geom_point() +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic() +
  coord_flip()
```

7. We are now going to calculate 95% binomial confidence intervals for the prevalence in each village. The formulae for the lower and upper bounds are given by:

- lower = p - 1.96 * sqrt(p * (1 - p) / n)
- upper = p + 1.96 * sqrt(p * (1 - p) / n)

where p is the proportion positive in each village and n is the sample size. Calculate these and save these in a data frame along with the village name and proportion positive. If the lower bound is below zero, set the lower bound to be zero.

```{r}
df_villages <- df %>% 
  group_by(village) %>% 
  summarise(
    p=mean(overall_sth_microscopy),
    lower=p-1.96 * sqrt(p * (1 - p) / n()),
    upper=p+1.96 * sqrt(p * (1 - p) / n())
  ) %>% 
  mutate(lower=if_else(lower<0, 0, lower))
```

8. Using the `geom_pointrange` command, draw point ranges which display the lower and upper bounds calculated in the previous question, and the positive proportion in each village.

Hint: `geom_pointrange` takes as arguments: `ymin`, `ymax` and `y`.

```{r}
df_villages %>% 
  ggplot(aes(x=village, y=p)) +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic()
```
9. We are now going to stratify the dataset by age, to explore how age impacts prevalence. To do so, run the following code (which assumes your dataset is called `df`).

```{r}
bins <- c(0, 2, 5, 10, 15, seq(20, 80, by = 10))
age_bins <- cut(df$age,
                breaks = bins,  
                right = FALSE,
                include.lowest = TRUE)
midpoints <- (head(bins, -1) + tail(bins, -1)) / 2
lookup <- tibble(
  age_bin=levels(age_bins),
  age_bin_mid=midpoints
)

df <- df %>% 
  mutate(age_bin=age_bins) %>% 
  left_join(lookup, by="age_bin") %>% 
  mutate(age_bin=as.factor(age_bin)) %>% 
  mutate(age_bin=fct_reorder(age_bin, age_bin_mid))
```

10. Calculate village-level prevalence by age_bin; then use ggplot to produce "age bin vs prevalence" point plots across all villages, each in separate facet panels; to do so, use: `+ facet_wrap(~village)`.

(Advanced: add in 95% confidence intervals.)

```{r}
df_villages <- df %>% 
  group_by(village, age_bin, age_bin_mid) %>% 
  summarise(
    p=mean(overall_sth_microscopy),
    lower=p-1.96 * sqrt(p * (1 - p) / n()),
    upper=p+1.96 * sqrt(p * (1 - p) / n())
  ) %>% 
  mutate(lower=if_else(lower<0, 0, lower))

df_villages %>% 
  ggplot(aes(x=age_bin, y=p)) +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic() +
  facet_wrap(~village)
```

11. Reduce the limit on the vertical axis to be only between 0 and 1.
```{r}
df_villages %>% 
  ggplot(aes(x=age_bin, y=p)) +
  geom_point() +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  scale_y_continuous(
    labels=scales::percent,
    limits = c(0, 1)
    ) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic() +
  facet_wrap(~village)
```

12. Rotate the horizontal axis labels by 90 degrees to make them more readable.
```{r}
df_villages %>% 
  ggplot(aes(x=age_bin, y=p)) +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic() +
  facet_wrap(~village) +
  theme(
    axis.text.x = element_text(angle=90)
  )
```

13. In a new plot, plot prevalence vs `age_bin_mid` (i.e. rather than `age_bin`). Use the following code oto add in a loess regression line (without a standard error):

`+ geom_smooth(se=FALSE)`

```{r}
df_villages %>% 
  ggplot(aes(x=age_bin_mid, y=p)) +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  scale_y_continuous(labels=scales::percent) +
  xlab("Village") +
  ylab("Prevalence, %") +
  theme_classic() +
  facet_wrap(~village) +
  geom_smooth(se=FALSE) +
  theme(
    axis.text.x = element_text(angle=90)
  )
```

